Threshold Phenomenon in a Nonlinear Heat Equation

### 2016-04-12 math_modeling/class/notes/stochastics.tex
number theory and modeling

@@

>>>>

\usepackage{tikz}
\usetikzlibrary{automata,positioning}

@@

\section*{Poisson Probabilities}

Suppose that $f(x,t)$ is the probability of getting $x$ successes during a time interval of length
$t$. Assume the prob of success during a very small time internal from $t$ to $t+\delta t$ is $\mu$,
the probability of more than one success during a time interval is negligible, and the probability
of a success during such a time interval does not depend on what happened prior to time $t$. The
Poisson distribution is given by
\[
f(x,t+\Delta t)=f(x,t)\cdot[1-\mu\Delta t]+f(x-1,t)\mu\Delta t = {e^{-\mu}\mu^{x}\over x!}
\]
The variable $x$ is indicative of the number of successes we want in the next time interval
$t+\delta t$. One example of the use of Poisson distributions is to calculate the probability that
an earthquake will occur again in a certain area.\\\\

\begin{quote}
\textit{\textbf{Example}}\\
Suppose that in southern California, 5 earthquakes occur every 100 years on average. What is the
probability that 2 earthquakes will occur in the next century?\\\\

The answer is very straightforward:
\[
P={e^{-5}5^{2}\over 2!}
\]
If we wanted to know the probability in the next 10 years, we can calculate
\[
P={e^{-0.5}0.5^{2}\over 2!}.
\]
\end{quote}

\section*{Markov Chains and Continuity}
We can form a relationship between Markov Chains and differential equations.
\begin{quote}
\textit{\textbf{Example}}\\
In the summer, an AC unit is in one of 3 states: (0) off, (1) low, or (2) high. While off,
transitions to low occur after an exponential time with an expected time of 3 minutes.
\begin{center}
\begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
    \node[state] (off) {off};
    \node[state] (low)[right=of off] {low};
    \node[state] (high)[right=of low] {high};
    \path[->]
    (off) edge[bend right,swap] node {$1/3$} (low)
    (low) edge[bend right,swap] node {$1/4$} (off)
          edge[bend right,swap] node {$1/4$} (high)
    (high) edge[bend right,swap] node {$1/3$} (low);
\end{tikzpicture}
\end{center}
We assign a $\lambda$ value to each state where $\lambda\in(0,\infty)$. The density function is
given by
\[
f(t)=\lambda e^{-\lambda t}
\]
for $t>0$. We can call the waiting time $T_{n}$, where $n$ is the state where we want to move. If
we have $T_{1}$ and $T_{2}$, the smaller $T$ is the state to which we move.
\end{quote}

\section*{Various Distributions}
\begin{center}
\bgroup
\def\arraystretch{3.5}
\begin{tabular}{|c | c | c | c|}
\hline
Binomial Distribution & $B(x;n,p)$ & $\displaystyle \binom{n}{x}p^{x}q^{n-x}$ & Discreet\\
\hline
Poisson Distribution & $\rho(x,\mu)$ & $\displaystyle {e^{-\mu}\mu^{x}\over x!}$ & Discreet\\
\hline
Geometric Distribution & $G(x;p)$ & $\displaystyle p(1-p)^{x-1}$ & Discreet, Memoryless\\
\hline
Bell Curve & $B(\sigma,x,\mu)$ & $\displaystyle \int_{a}^{b}{1\over \sigma\sqrt{2\pi}}e^{-0.5\left({x-\mu\over\sigma}\right)^{2}}\dx$ & Continuous\\
\hline
Exponential Distribution & $E(t,\theta)$ & $\displaystyle\int_{a}^{b}{1\over\theta}e^{-t/\theta}\dt$ & Continuous, Memoryless\\
\hline
\end{tabular}
\egroup
\end{center}

### 2016-04-12 math_modeling/hw/hw6/hw6.tex
finished pres for now
number theory and modeling

\begin{minipage}{0.5\textwidth}
\[
\begin{array}{c | c c c}
& \text{L} & \text{C} & \text{N}\\
\hline
\text{L} & .6 & .2 & .2\\
\text{C} & .2 & .8 & 0\\
\text{N} & .15 & .1 & .75
\end{array}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{center}
\begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
   \node[state] (q_0) {L};
   \node[state] (q_1) [right=of q_0] {C};
   \node[state] (q_2) [right=of q_1] {N};
   % \node[state] (q_2) [below right=of q_0] {$q_2$};
   % \node[state,accepting](q_3) [below right=of q_1] {$q_3$};
    \path[->]
    (q_0)   edge [bend left] node {0.2} (q_1)
            edge [bend left=70] node {0.2} (q_2)
            edge [loop left] node {0.6} ()
    (q_1)   edge [bend left] node {0.2} (q_0)
            edge [loop above] node {0.8} ()
    (q_2)   edge [bend left=70] node {.15} (q_0)
            edge [bend left] node {.1} (q_1)
            edge [loop right] node {.75} ();
\end{tikzpicture}
\end{center}
\end{minipage}
\end{enumerate}

@@

\item Find the probability that a woman whose birthweight was average has a granddaughter with an
average birthweight.
\[
\begin{array}{c | c c}
& \text{Average} & \text{Not Average}\\
\hline
\text{Average} & .5 & .5\\
\text{Not Average} & .5 & .5
\end{array}
\]
\begin{quote}
The probability that a granddaughter will have average weight is $(.5)(.5)=.25$.
\end{quote}

@@

\item Write out inductive proofs for Theorems 2 and 3.

\item Use matrix multiplication to solve Exercise 8.
\[
\left(
\begin{array}{c c}
.5 & .5\\
.5 & .5
\end{array}
\right)
\cdot
\left(
\begin{array}{c c}
.5 & .5\\
.5 & .5
\end{array}
\right)
=
\]

\item Abigail bought a toy with her allowance this week (see Exercise 2). Find the probability that
she will buy a toy 4 weeks from now.

\item Find the distribution of birth weights (Example 1) after one generation if the initial
probability distribution is (.4, .3, .3).

\item Suppose the distribution of birth weights of a generation of daughters is
$p^{(1)}=7(0.31,0.45,0.24)$. Can you find the distribution of birth weights of the mothers?

\item Use the state diagram in Fig. 11.3 to find the probability that the process reaches state
$S_{3}$ in two steps if it starts in state $S_{1}$.

\item Let $P$ be the transition matrix of Eq. (4). Compute $P^{2}$. What can you say about the
first column of $P^{3}$ and $P^{n}$?

>>>>

\begin{minipage}{0.5\textwidth}
\[
\begin{array}{c | c c c}
& \text{L} & \text{C} & \text{N}\\
\hline
\text{L} & .6 & .2 & .2\\
\text{C} & .2 & .8 & 0\\
\text{N} & .15 & .1 & .75
\end{array}
\]
\end{minipage}
\begin{minipage}{0.5\textwidth}
\begin{center}
\begin{tikzpicture}%[shorten >=1pt,node distance=15cm,on grid,auto]
   \node[state] (q_0) {L};
   \node[state] (q_1) [right=of q_0] {C};
   \node[state] (q_2) [right=of q_1] {N};
   % \node[state] (q_2) [below right=of q_0] {$q_2$};
   % \node[state,accepting](q_3) [below right=of q_1] {$q_3$};
    \path[->]
    (q_0)   edge [bend left] node {0.2} (q_1)
            edge [bend left=70] node {0.2} (q_2)
            edge [loop left] node {0.6} ()
    (q_1)   edge [bend left] node {0.2} (q_0)
            edge [loop above] node {0.8} ()
    (q_2)   edge [bend left=70] node {.15} (q_0)
            edge [bend left] node {.1} (q_1)
            edge [loop right] node {.75} ();
\end{tikzpicture}
\end{center}
\end{minipage}
\end{enumerate}

@@

\item Find the probability that a woman whose birthweight was average has a granddaughter with an
average birthweight.
\[
\begin{array}{c | c c}
& \text{Average} & \text{Not Average}\\
\hline
\text{Average} & .5 & .5\\
\text{Not Average} & .5 & .5
\end{array}
\]
\begin{quote}
The probability that a granddaughter will have average weight is $(.5)(.5) + (.5)(.5)=.5$.
\end{quote}

@@

\item Write out inductive proofs for Theorems 2 and 3.
\begin{quote}
\textbf{Theorem 2}: From the text, we know
\[
p^{(0)}=(p_{1}^{(0)},p_{2}^{(0)},\dots p_{r}^{(0)})=
(p_{1},p_{2},\dots p_{r})=p^{0}\cdot p.
\]
We can assume that $p^{(n)}=p^{n}$, and we want to show that $p^{(n+1)}=p^{n+1}$. We know that
\begin{align*}
p^{(n+1)}&=(p_{1}^{(n+1)},p_{2}^{(n+1)},\dots p_{r}^{(n+1)})\\
&= (p_{1}^{(n)}p_{1}^{(1)},p_{2}^{(n)}p_{2}^{(1)},\dots p_{r}^{(n)}p_{r}^{(1)})\\
&= p^{n+1}.
\end{align*}
\textbf{Theorem 3}: We know that $p^{(n)}=p^{(0)}p^{n}$. We want to show $p^{(n+1)}=p^{(0)}p^{n+1}$.
From theorem 2, we have
\begin{align*}
p^{(n+1)}&=(p_{1}^{(n+1)},p_{2}^{(n+1)},\dots p_{r}^{(n+1)})\\
&= (p_{1}^{(n)}p_{1}^{(0)},p_{2}^{(n)}p_{2}^{(0)},\dots p_{r}^{(n)}p_{r}^{(0)})\\
&= p^{n+1}p^{(0)}.
\end{align*}
\end{quote}

\item Use matrix multiplication to solve Exercise 8.
\begin{quote}
\[
\left(
\begin{array}{c c}
.5 & .5\\
.5 & .5
\end{array}
\right)
\cdot
\left(
\begin{array}{c c}
.5 & .5\\
.5 & .5
\end{array}
\right)
=
\left(
\begin{array}{c c}
.5 & .5\\
.5 & .5
\end{array}
\right)
\]
Therefore, the probability is .5.
\end{quote}

\item Abigail bought a toy with her allowance this week (see Exercise 2). Find the probability that
she will buy a toy 4 weeks from now.
\begin{quote}
The probability that she will buy a toy can be found by raising the matrix to the fourth power.
\[
\left(
\begin{array}{c c}
.4 & .6\\
.8 & .2
\end{array}
\right)^{4}
=
\left(
\begin{array}{c c}
.5824 & .4176\\
.5568 & .4432
\end{array}
\right).
\]
Therefore, the probability is $0.4432$.
\end{quote}

\item Find the distribution of birth weights (Example 1) after one generation if the initial
probability distribution is $(.4, .3, .3)$.
\[
(.4,.3,.3)\cdot
\left(
\begin{array}{c c c}
.5 & .45 & .05\\
.25 & .5 & .25\\
.3 & .3 & .4
\end{array}
\right)
= (.365,.42,.215).
\]

\item Suppose the distribution of birth weights of a generation of daughters is
$p^{(1)}=7(0.31,0.45,0.24)$. Can you find the distribution of birth weights of the mothers?
\begin{quote}
We need to solve the system
\[
(x,y,z)
\left(
\begin{array}{c c c}
.5 & .45 & .05\\
.25 & .5 & .25\\
.3 & .3 & .4
\end{array}
\right)
= 7(.31,.45,.24).
\]
Doing so yields
\[
(x,y,z)=(1.4,4.2,1.4).
\]
\end{quote}

\item Use the state diagram in Fig. 11.3 to find the probability that the process reaches state
$S_{3}$ in two steps if it starts in state $S_{1}$.
\begin{quote}
To move from $S_{1}$ to $S_{2}$ has a probability of $a$, and from $S_{2}$ to $S_{3}$ has a
probability $(1-b)$, so the total probability is $a(1-b)$.
\end{quote}

\item Let $P$ be the transition matrix of Eq. (4). Compute $P^{2}$. What can you say about the
first column of $P^{3}$ and $P^{n}$?
\begin{quote}
\[
P^{2}=
\left(
\begin{array}{c c c}
0 & a^2 & (1-a)^2\\
0 & b^2 & (1-b)^2\\
0 & c^2 & (1-c)^2
\end{array}
\right).
\]
The first column of $P^{n}$ will always be 0.
\end{quote}

### 2016-04-12 pde_two/threshold.tex
number theory and modeling

\subsection{Solution to the Steady State Problem}
We are attempting to solve the ordinary differential equation
\[
u''=-u^{3}.
\]
Multiplying both sides of the equation gives us
\begin{align*}
u''u'&=-u'u^{3}\\
\int u''u'&=\int -u'u^{3}\\
{1\over2}(u')^{2}&= -{1\over4}u^{4}+c\\
2(u')^{2}&=-u^{4}+c_{1}.
\end{align*}
This first order equation does not lend itself to separation of variables as readily as we would like
(although it is possible if we want to deal with imaginary numbers). Instead, we will use a technique
called the \textit{shooting method} in order to determine a solution. The shooting method provides
us with a way to relax boundary conditions into initial conditions. Recall that $(5)$ provides us
with boundary conditions that arose from the steady state of $(1)$. We will leave the first boundary
condition alone and change the second boundary condition so that

>>>>

\subsection{Solution to the Steady State Problem}
We are attempting to solve the ordinary differential equation
\[
u''=-u^{3}.
\]
Multiplying both sides of the equation gives us
\begin{align*}
u''u'&=-u'u^{3}\\
\int u''u'&=\int -u'u^{3}\\
{1\over2}(u')^{2}&= -{1\over4}u^{4}+c\\
2(u')^{2}&=-u^{4}+c_{1}.
\end{align*}
This first order equation does not lend itself to separation of variables as readily as we would like
(although it is possible if we want to deal with imaginary numbers). Instead, we will use a technique
called the \textit{shooting method} in order to determine a solution. The shooting method provides
us with a way to relax boundary conditions into initial conditions. Recall that $(5)$ provides us
with boundary conditions that arose from the steady state of $(1)$. We will leave the first boundary
condition alone and change the second boundary condition so that $2u'(0)^{2}=a^{4}$. This will allow
us to factor the resulting term:
\begin{align*}
u'(x)^2&={1\over2}(a^{4}-u^{4})\\
{\du\over\dx}&= \sqrt{{1\over2}(a^{2}+u^{2})(a^{2}-u^{2})}\\
{\dx\over\du}&= {1\over\sqrt{{1\over2}(a^{2}+u^{2})(a^{2}-u^{2})}}.
\end{align*}
Integrating both sides and substituting $s$ for $u$ gives us
\[x=\int_{0}^{u}{1\over\sqrt{{1\over2}(a^{2}+s^{2})(a^{2}-s^{2})}}\mathop{ds}.\]
We can now make another substitution for $s$ in terms of $t$. Letting
\[
s=\sqrt{{a^2t^2\over2(1-t^2/2)}}
\]
yields
\[
a^2+s^2={2a^2\over2(1-t^2/2)}\qquad\qquad\text{and}\qquad\qquad
a^2-s^2={a^2(1-t^2)\over(1-t^2/2)}.
\]
We also have
\[
\mathop{ds}={a(t-1)(t-2)\over2\sqrt{2}(1-t^2/2)^{3/2}}\dt,
\]
which gives us the new integral
\[
ax=\int_{0}^{\sqrt{\frac{\displaystyle2u}{\displaystyle a^2+u^2}}}{1\over\sqrt{(1-t^2)(1-t^2)}}.
\]
We may notice that this integral is one of the ways to define $sn(x,k)$. Rewriting it in terms
of our Jacobi elliptic function gives us
\[
x=\int_{0}^{sn(x,k)}{1\over\sqrt{(1-t^2)(1-t^2/2)}}.
\]
